Blum axioms and nondeterministic computation of functions

### 2016-04-20 paper.tex
paper.tex: Remove some superfulous information.

But nondeterministic deciders,
although computationally equivalent to deterministic ones,
are apparently exponentially faster.
That is, while we have linear-time non-deterministic algorithms for Boolean satisfiability,
the best known deterministic algorithms are still exponential in the worst case.
Nondeterminism seems special in this sense;
although other extensions (multiple tapes and doubly-infinite tapes)
can be simulated under just a polynomial slowdown,
the simulation of nondeterminism seems to require exponential slowdown.%
\footnote{
    Simulating nondeterminism under a polynomial slowdown would show that $\P = \NP$;
    showing that the slowdown must be exponential would show that $\P \neq \NP$.
    Although most researchers expect the later to be the case~\cite[p.~54]{Gasarch2012},
    as Papadimitriou noted~\cite[p.~412]{Papadimitriou1994},
    in the absence of a proof that $\P \neq \NP$
    we should not be too emphatic in stating the simulation \emph{require}
    (as opposed to \emph{seemingly require}) exponential slowdown.
}
Therefore,
our definition must exhibit this apparent exponential speedup
when computing, for instance, the characteristic function\footnotemark{} of $\SAT$.
\footnotetext{
    The characteristic function of a set $A \subseteq \Sigma^*$
    is the function $f$ defined by
    \begin{equation*}
        f(x) = \begin{cases}
            1, & \text{if $x \in A$}. \\
            0, & \text{if $x \notin A$}.
        \end{cases}
    \end{equation*}
}
This is the ``non-deterministic'' part.

@@

Therefore,
these encodings provide a way to associate a string
(which is a finite, manipulable object)
with a partial recursive function
(which is an infinite, mathematical, ``untouchable'' object).
Encodings are \emph{enumerations} of all partial recursive function.

In some sense,
we are going the opposite direction:
instead of associating functions with strings,
we want to associate machines with functions.

@@

The concept of \emph{acceptable Gödel numbering}
(\cite[p.~41]{Rogers1987},~\cite[p.~324]{Blum1967})
encompasses the existence of universal machines and a little more.
We will use it as our reasonableness criteria to our definitions.%

@@

Condition~\ref{surjectiveness}
guarantees that the image of~$\phi$ is all of~$\mathcal P$,
so that the numbering neither ``forgets'' a partial recursive function
nor generate a function that is not partial recursive.

Condition~\ref{universal-tm} is the universal Turing machine theorem.

@@

\begin{example}
    We can also think of less orthodox numberings.

    As two-stack deterministic pushdown automata are equivalent to Turing machines%
    ~\cite[p.~172]{HopcroftUllman1979},
    any encoding of these kind of automaton yields acceptable Gödel numberings.

    Unrestricted grammars can also simulate Turing machines%
    ~\cite[p.~221]{HopcroftUllman1979};
    so if a grammar $G$ generate pairs $(x, y)$,
    define the corresponding partial recursive function $f$
    by $f(x) = y$ if,
    among all the pairs whose first element is $x$,
    the pair $(x, y)$ is the one with smallest number of productions
    (leave $f$ undefined if there are ties or no pairs with $x$ as first element).
    (Choosing the pair $(x, y)$ with the smaller number of productions
    allows a Turing machine to simulate this function
    by doing a breadth-first search in the grammar,
    guaranteeing the function is partial recursive.)

    And, as an off-line Turing machine
    (a machine with a special read-only tape for its input)
    can simulate any Turing machine without ever writing a $0$ over a~$1$
    \cite[p.~174]{HopcroftUllman1979},
    we can associate an encoding of a off-line Turing machine
    with the function that counts the number of times
    it writes a $0$ over an $1$ before halting
    (the number can be converted to binary and have its most significant digit removed
    to generate all possible binary strings).
    (The machine can choose how many times it will write a $0$ over an $1$,
    thus we can generate any partial recursive function by manipulating this choice.)
\end{example}

There are several interesting properties we can show about acceptable Gödel numberings.
For instance,
if $\phi$ and $\phi'$ are any two such numberings,
there is a total recursive function that converts programs for $\phi$
into programs for $\phi'$;
that is, there is a total recursive function $f : \{0, 1\}^* \to \{0, 1\}^*$
such that, for every program $w$,
\begin{equation*}
    \phi_w = \phi'_{f(w)}.
\end{equation*}
To show this,
we can use the $\sigma$ function for $\phi'$
to fix the input program as the first variable of the $U$ function for $\phi$.
With a bit more work,
it is possible to make $f$ bijective \cite[p.~25]{Soare1987}.

Another important theorem is the recursion theorem
\cite[p.~181]{Rogers1987}.
It states that, if $\phi$ is any acceptable Gödel numbering
and $f$ is any total recursive function,
then there is some program $w_0$ such that $\phi_{w_0} = \phi_{f(w_0)}$.
That is,
if $f$ is any systematic transformation on programs,
there is a program $w$ (a fixed point for $f$) whose meaning under $\phi$ is unchanged.
The recursion theorem can be used,
for example, to show the existence of quines
(programs whose output are their own source codes)
in any Turing-complete programming language:
choose $f$ to be the function that, given a program $w$,
returns another program $f(w)$ that prints the string $w$ when run.
(For Turing machines, for instance,
we can encode the bits of $w$ in the transition table of $w$.)
Then, by the recursion theorem,
there is some program $w_0$ that is equivalent to its transformed version $f(w_0)$;
thus, $w_0$ already writes the string $w_0$, its own source code.
Therefore, any programming language has quines \cite[p.~227]{Kozen2006}.

@@

Note that the complexity measure must be subordinated to the machine model.
For instance,
if we are talking about pushdown automata with two stacks,
we could restrict the machine to operate on one stack each time
(that is, each move must alter at most one stack)
and measure the number of times the automaton switches its ``active'' stack.
This complexity measure makes no sense for Turing machines,
which have no stacks.

Therefore,
we must have an acceptable Gödel numbering $\phi$
to discuss complexity.
We can think that a complexity measure is a way of associating
each program $w$ and each input $x$
with a number $n$,
which is the amount of the measured resource it takes to obtain $\phi_w(x)$
using the program $w$.

As we did with the machine encodings,
we will impose some restrictions on what can be a complexity measure
to be able to manipulate it (at least indirectly).
In our case,
we will use Blum axioms~\cite[p.~324]{Blum1967}.

@@

\begin{example}
    As with acceptable Gödel numberings,
    we can also measure less usual resources.
    For example, we can satisfy the Blum axioms by
    counting the number of times a machine changes a symbol in the tape,
    the number of times the tape head moves right,
    or the number of times a two-stack pushdown automata
    pops a symbol from the second stack.
\end{example}

@@

\begin{proof}
    Note that the domain in which $f$ is defined is exactly
    the complement (under $\Sigma^*$) of the halting problem.
    If $f$ were computable,
    there would be a deterministic machine that could compute it;
    we could, then, modify the machine to accept all inputs in the domain of $f$
    (all inputs that have some output through $f$),
    thus recognizing (albeit not deciding) the complement of the halting problem.
    But, as the halting problem itself is computable but not decidable,
    its complement cannot also be --- otherwise both would be decidable,
    a contradiction.

@@

    Thus, under Hopcroft and Ullman's definition,
    this machine computes the function $f$,
    which we already shown to be noncomputable.
\end{proof}

@@

Therefore, with this restriction,
we lose the apparent exponential speed up in computation time
we have when using nondeterminism.
We thus will analyze the following models
to see whether they also suffer from this problem.

@@

\begin{definition}[Goldreich's definition]
    Let $\bot$ be some special symbol not in $\{0, 1\}^*$.
    (This symbol will represents ``don't know'')
    A nondeterministic machine $M$ computes the function $f$ if,
    when processing the input $x$,
    both the following conditions hold \cite[p.~168]{Goldreich2008}:
    \begin{itemize}
        \item Every branch of $M$ halts
            and outputs either $f(x)$ or $\bot$.
        \item At least one branch of $M$ halts with $f(x)$ on the tape.
    \end{itemize}
\end{definition}

@@

At least partially,
we can compute the characteristic function of the satisfiability problem
in linear time using this model.
The machine $M$ that will do the calculation will guess an assignment
and write $1$ if it is satisfying and write $\bot$ if it isn't.
This computes in linear time the function $f$, defined by
\begin{equation*}
    f(\varphi) = \begin{cases}
        1, & \text{if $\varphi$ is satisfiable.} \\
        \text{undefined}, & \text{if $\varphi$ is unsatisfiable.} \\
    \end{cases}
\end{equation*}
However,
this function falls short of being exactly $\SAT$'s characteristic function:
$f(\varphi)$ ought to be zero for unsatisfiable formulas $\varphi$.
Although we cannot prove a strong assertion as
``if this works then $\P = \NP$'',
we can come close.

@@

\subsubsection{Extension $\NP$-completeness}
\label{sec:np-completeness-extension}

@@

Krentel's and Valiant's definitions
(which were discovered later in the project),
although not created specifically to associate nondeterministic machine with functions,
provides a more elegant generalization of $\NP$ and $\NP$-completeness;
this paper summarize all these definitions
under the light of Gödel numberings and Blum axioms.

>>>>

But nondeterministic deciders,
although computationally equivalent to deterministic ones,
are apparently exponentially faster.
That is, while we have linear-time non-deterministic algorithms for Boolean satisfiability,
the best known deterministic algorithms are still exponential in the worst case.
Nondeterminism seems special in this sense;
although other extensions (multiple tapes and doubly-infinite tapes)
can be simulated under just a polynomial slowdown,
the simulation of nondeterminism seems to require exponential slowdown.%
\footnote{
    Simulating nondeterminism under a polynomial slowdown would show that $\P = \NP$;
    showing that the slowdown must be exponential would show that $\P \neq \NP$.
    Although most researchers expect the later to be the case~\cite[p.~54]{Gasarch2012},
    as Papadimitriou noted~\cite[p.~412]{Papadimitriou1994},
    in the absence of a proof that $\P \neq \NP$
    we should not be too emphatic in stating the simulation \emph{require}
    (as opposed to \emph{seemingly require}) exponential slowdown.
}
Therefore,
our definition must exhibit this apparent exponential speedup
when computing, for instance, the characteristic function\footnotemark{} of $\SAT$.
\footnotetext{
    The characteristic function of a set $A$
    is the function $1_A$ defined by
    \begin{equation*}
        1_A(x) = \begin{cases}
            1, & \text{if $x \in A$}. \\
            0, & \text{if $x \notin A$}.
        \end{cases}
    \end{equation*}
}
This is the ``non-deterministic'' part.

@@

Therefore,
these encodings provide a way to associate a string
(which is a finite, manipulable object)
with a partial recursive function
(which is an infinite, mathematical, ``untouchable'' object).
Encodings are \emph{enumerations} of all partial recursive functions.

@@

The concept of \emph{acceptable Gödel numbering}
(\cite[p.~41]{Rogers1987},~\cite[p.~324]{Blum1967})
encompasses the existence of universal machines and a little more.
We will use it as our reasonableness criteria to our definitions.

@@

Condition~\ref{surjectiveness}
guarantees that the image of~$\phi$ is all of~$\mathcal P$,
so that the numbering neither ``forgets'' a partial recursive function
nor generate a function that is not partial recursive.
Condition~\ref{universal-tm} is the universal Turing machine theorem.

@@

One important theorem that can be proven using acceptable Gödel numberings alone
is the recursion theorem
\cite[p.~181]{Rogers1987}.
It states that, if $\phi$ is any acceptable Gödel numbering
and $f$ is any total recursive function,
then there is some program $w_0$ such that $\phi_{w_0} = \phi_{f(w_0)}$.
That is,
if $f$ is any systematic transformation on programs,
there is a program $w$ (a fixed point for $f$) whose meaning under $\phi$ is unchanged.
The recursion theorem can be used,
for example, to show the existence of quines
(programs whose output are their own source codes)
in any Turing-complete programming language:
choose $f$ to be the function that, given a program $w$,
returns another program $f(w)$ that prints the string $w$ when run.
(For Turing machines, for instance,
we can encode the bits of $w$ in the transition table of $w$.)
Then, by the recursion theorem,
there is some program $w_0$ that is equivalent to its transformed version $f(w_0)$;
thus, $w_0$ already writes the string $w_0$, its own source code.
Therefore, any programming language has quines \cite[p.~227]{Kozen2006}.

@@

Note that the complexity measure must be subordinated to the machine model.
For instance,
in an assembly language,
we could measure the number of \texttt{jump} instructions executed during execution.
This complexity measure makes no sense for Turing machines,
which have no stacks.

Therefore,
we must have an acceptable Gödel numbering $\phi$
to discuss complexity.
As we did with the machine encodings,
we will impose some restrictions on what can be a complexity measure
to be able to manipulate it (at least indirectly).
In our case,
we will use Blum axioms~\cite[p.~324]{Blum1967}.

@@

@@

\begin{proof}
    If $f$ were computable,
    we could take a deterministic machine $M$ that computes it
    and replacing halting
    (which only happen if $M$ would return $1$)
    with acceptance.
    This new machine would recognize the complement (under $\{0, 1\}^*$)
    of the halting problem,
    a contradiction.

@@

    Thus, under Hopcroft and Ullman's definition,
    this machine computes the noncomputable function $f$,
\end{proof}

@@

Therefore, with this restriction,
we lose the apparent exponential speed up in computation time
we have when using nondeterminism.

@@

\begin{definition}[Goldreich's definition]
    Let $\bot$ be some special symbol not in $\{0, 1\}^*$.
    (This symbol will represents ``don't know''.)
    A nondeterministic machine $M$ computes the function $f$ if,
    when processing the input $x$,
    both the following conditions hold \cite[p.~168]{Goldreich2008}:
    \begin{itemize}
        \item Every branch of $M$ halts
            and outputs either $f(x)$ or $\bot$.
        \item At least one branch of $M$ halts with $f(x)$ on the tape.
    \end{itemize}
\end{definition}

@@

Take, for instance,
the characteristic function of the satisfiability problem.
We can create a machine that guesses an assignment
and write $1$ if it is satisfying and write $\bot$ if it isn't.
This computes in linear time the function $f$, defined by
\begin{equation*}
    f(\varphi) = \begin{cases}
        1, & \text{if $\varphi$ is satisfiable.} \\
        \text{undefined}, & \text{if $\varphi$ is unsatisfiable.} \\
    \end{cases}
\end{equation*}
However,
this function falls short of being exactly $\SAT$'s characteristic function:
$f(\varphi)$ ought to be zero for unsatisfiable formulas $\varphi$.
Although we cannot prove a strong assertion as
``if this works then $\P = \NP$'',
we can come close.

@@

\subsubsection{Extending $\NP$-completeness}
\label{sec:np-completeness-extension}

@@

Krentel's and Valiant's definitions
(which were discovered later in the project),
although not created specifically to associate nondeterministic machine with functions,
provides a more elegant generalization of $\NP$ and $\NP$-completeness;
this paper summarize all these findings
under the light of Gödel numberings and Blum axioms.
